# Foundations for Inference


```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=2.5,
                      fig.width=5, 
                      fig.align = "center", cache=TRUE)
```

## 1/23/20 Agenda {#Jan23}
1. Example: gender discrimination
2. `infer` again
3. Hypothesis testing structure




## Example: Gender Discrimination {#ex:gend}

> We consider a study investigating gender discrimination in the 1970s, which is set in the context of personnel decisions within a bank.^[Rosen B and Jerdee T. 1974. Influence of sex role stereotypes on personnel decisions. Journal of Applied Psychology 59(1):9-14.] The research question we hope to answer is, "Are females discriminated against in promotion decisions made by male managers?"

> The participants in this study were 48 male bank supervisors attending a management institute at the University of North Carolina in 1972. They were asked to assume the role of the personnel director of a bank and were given a personnel file to judge whether the person should be promoted to a branch manager position. The files given to the participants were identical, except that half of them indicated the candidate was male and the other half indicated the candidate was female. These files were randomly assigned to the subjects.

> For each supervisor we recorded the gender associated with the assigned file and the promotion decision. Using the results of the study summarized in Table 2.1, we would like to evaluate if females are unfairly discriminated against in promotion decisions. In this study, a smaller proportion of females are promoted than males (0.583 versus 0.875), but it is unclear whether the difference provides convincing evidence that females are unfairly discriminated against.   (@isrs, pg 61) 

|   |        |          | decision     |       |
|---|--------|----------|--------------|-------|
|   |        | promoted | not promoted | total |
|   | male   | 21       | 3            | 24    |
|   | female | 14       | 10           | 24    |
|   | total  | 35       | 13           | 48    |


#### Always Ask {-}

* What are the observational units?
    - supervisor
* What are the variables?  What type of variables?
    - (1) whether the resume was male or female (categorical)
    - (2) decision to promote or not promote (categorical)
* What is the statistic?
    - $\hat{p}_m - \hat{p}_f$ = 21/24 - 14/24 = 0.292  (the difference between the proportion of men who were promoted and the proportion of women who were promoted)
* What is the parameter?
    - $p_m - p_f$ = the true difference in the probability of a man being promoted minus the probability of a woman being promoted.
    
#### Hypotheses {-}

H0: Null hypothesis. The variables gender and decision are independent. They have no relationship, and therefore any observed difference between the proportion of males and females who were promoted is due to chance.

HA: Alternative hypothesis. The variables gender and decision are not independent. Any observed difference between the proportion of males and females who were promoted is **not** due to chance.

#### Computation {-}

```{r echo=TRUE, eval=TRUE}
library(infer)

# to control the randomness
set.seed(47)

# first create a data frame with the discrimination data
discrim <- data.frame(gender = c(rep("male", 24), rep("female", 24)),
                      decision = c(rep("promote", 21), rep("not", 3), 
                                   rep("promote", 14), rep("not", 10)))

discrim %>% head()


# then find the difference in proportion who are promoted
(diff_obs <- discrim %>%
    specify(decision ~ gender, success = "promote") %>%
    calculate(stat = "diff in props", order = c("male", "female")) )
    

# now apply the infer framework to get the null differences in proportions
null_discrim <- discrim %>%
  specify(decision ~ gender, success = "promote") %>%
  hypothesize(null = "independence") %>%
  generate(reps = 10000, type = "permute") %>%
  calculate(stat = "diff in props", order = c("male", "female"))

# then visualize the null sampling distribution & p-value
visualize(null_discrim, bins = 10) +
  shade_p_value(obs_stat = diff_obs, direction = "greater")

# calculate the actual p-value
null_discrim %>%
  get_p_value(obs_stat = diff_obs, direction = "greater")
```

#### Logic for what we believe {-}

1. We know that the study was an experiment, so there should be no **systematic** differences between the group who received "male" applications and "female" applications.

2. We've ruled out random chance as the reason for the huge difference in proportions.  (We reject the null hypothesis.)  if we lived in the null reality, we'd only see data like these about 2.5% of the time.

3. We conclude that gender and decision are not independent.  That is, knowing the gender changes the probability of promotion.



## Structure of Hypothesis testing

### Hypotheses

* **Hypothesis Testing** compares data to the expectation of a specific null hypothesis.  If the data are unusual, assuming that the null hypothesis is true, then the null hypothesis is rejected.  

* The **Null Hypothesis**, $H_0$, is a specific statement about a population made for the purposes of argument.  A good null hypothesis is a statement that would be interesting to reject. 

* The **Alternative Hypothesis**, $H_A$, is a specific statement about a population that is in the researcher's interest to demonstrate.  Typically, the alternative hypothesis contains all the values of the population that are not included in the null hypothesis. 

* In a **two-sided** (or two-tailed) test, the alternative hypothesis includes values on both sides of the value specified by the null hypothesis.

* In a **one-sided** (or one-tailed) test, the alternative hypothesis includes parameter values on only one side of the value specified by the null hypothesis. $H_0$ is rejected only if the data depart from it in the direction stated by $H_A$.


### Other pieces of the process

* A  **statistic** is a numerical measurement we get from the sample, a function of the data. [Also sometimes called an **estimate**.]

* A  **parameter** is a numerical measurement of the population.  We never know the true value of the parameter.

* The **test statistic** is a quantity calculated from the data that is used to evaluate how compatible the data are with the result expected under the null hypothesis.

* The **null distribution** is the sampling distribution of outcomes for a test statistic under the assumption that the null hypothesis is true. 

* The **p-value** is the probability of obtaining the data (or data showing as great or greater difference from the null hypothesis) if the null hypothesis is true.  *The p-value is a number calculated from the dataset.*


#### Examples of Hypotheses {-}
Identify whether each of the following statements is more appropriate as the null hypothesis or as the alternative hypothesis in a test:

* The number of hours preschool children spend watching TV affects how they behave with other children when at day care.  *Alternative*

* Most genetic mutations are deleterious.  *Alternative*

* A diet of fast foods has no effect on liver function.  *Null*

* Cigarette smoking influences risk of suicide.  *Alternative*

* Growth rates of forest trees are unaffected by increases in carbon dioxide levels in the atmosphere.  *Null*

* The number of hours that grade-school children spend doing homework predicts their future success on standardized tests.  *Alternative*

* King cheetahs on average run the same speed as standard spotted cheetahs. *Null*

* The risk of facial clefts is equal for babies born to mothers who take folic acid supplements compared with those from mothers who do not.  *Null*

* The mean length of African elephant tusks has changed over the last 100 years.  *Alternative*

* Caffeine intake during pregnancy affects mean birth weight.  *Alternative*

#### What is an Alternative Hypothesis? {-}

Consider the brief video from the movie Slacker, an early movie by Richard Linklater (director of Boyhood, School of Rock, Before Sunrise, etc.). You can view the video here from starting at 2:22 and ending at 4:30:  https://www.youtube.com/watch?v=b-U_I1DCGEY

In the video, a rider in the back of a taxi (played by Linklater himself) muses about alternate realities that could have happened as he arrived in Austin on the bus. What if instead of taking a taxi, he had found a ride with a woman at the bus station? He could have take a different road into a different alternate reality, and in that reality his current reality would be an alternate reality. And so on.

What is the point?  Why did we see the video?  How does it relate the to the material from class?  What is the relationship to sampling distributions?


###  All together:  structure of a hypothesis test

* decide on a research question (which will determine the test)
* collect data, **specify** the variables of interest
* state the null (and alternative) **hypothesis** values (often statements about parameters)
   - the null claim is the science we want to reject
   - the alternative claim is the science we want to prove
* **generate** a (null) sampling distribution to describe the variability of the statistic that was **calculated** along the way
* **visualize** the distribution of the statistics under the null model
* **get_p_value** to measure the consistency of the observed statistic and the possible values of the statistic under the null model
* make a conclusion using words that describe the research setting

## 1/28/20 Agenda {#Jan28}
1. Central Limit Theorem
2. Mathematical approximation for the distribution of one sample proportion

## Normal Model


### Central Limit Therm {#CLT}


#### Example: Reese's Pieces {-}

As with many of the examples, the Reese's Pieces example comes from @iscam.  The example focuses on how the samples of orange Reese's Pieces vary from sample to sample.  Today we aren't particularly interested in a specific research question, instead we are trying to understand the details of the model which describes how $\hat{p}$ varies from sample to sample.  [Spoiler: the distribution is going to look like a bell!  and the mathematical model which describes the variability is called the normal distribution.]

Notes from the applet: http://www.rossmanchance.com/applets/OneProp/OneProp.htm?candy=1

* How does the sampling distribution change as a function of $p$ and $n$?
* When a normal distribution is placed on top of the empirical (computational) distribution, does it fit well?




A *sampling distribution* is the probability distribution of all possible values of the *statistic* in all possible samples of the same size from the same population.  Note: increasing the sample size reduces the spread of the sampling distribution of a statistic (i.e., increases the precision).

##### Normal Probability Curve {-}

* symmetric
* bell-shaped
* centered at $\mu$
* $\sigma$ shows the point of inflection
* draw a picture **every** time you start a normal problem!

The Central Limit Theorem

> **The Central Limit Theorem** says that the sampling distribution of an  *average* will have a bell shaped distribution if $n$ is big enough.


The sampling distribution of $\hat{p} = X/n$ can be thought of as taking lots of random samples from a population, calculating $\hat{p}$, and creating a histogram.  We can easily calculate what we'd expect from that sampling distribution if we know $p$, the true population proportion.  

Because $\hat{p}$ is actually an average, the sampling distribution of $\hat{p}$ can be described by a normal distribution (as long as $n$ is big enough).

\begin{eqnarray*}
\hat{p} &=& \frac{X}{n}\\
SD(\hat{p}) = \sigma_{\hat{p}} &=& \sqrt{\frac{p (1-p)}{n}}\\
SE(\hat{p}) &=& \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}\\
\hat{p} &\sim& N\bigg(p, \sqrt{\frac{p(1-p)}{n}} \bigg) \ \ \ \ \ \mbox{ (if the sample size is large enough)}\\
\end{eqnarray*}


Notice the slight difference between $SD$ (uses $p$) and $SE$ (uses $\hat{p}$).  We won't make a big deal of the difference here (and indeed, your book calls both equations $SD$).  We would expect 95% of our $\hat{p}$ values to be within 2 standard deviations of the mean.  That is, 95% of $\hat{p}$ are:
\begin{eqnarray*}
p \pm 2 \sqrt{\frac{p(1-p)}{n}}
\end{eqnarray*}
Or put differently, when referring to a randomly selected $\hat{p}$,
\begin{eqnarray*}
P\bigg( - 2 \sqrt{\frac{p(1-p)}{n}} \leq \hat{p} - p \leq 2 \sqrt{\frac{p(1-p)}{n}}\bigg) = 0.95\\
P\bigg(\hat{p} - 2 \sqrt{\frac{p(1-p)}{n}} \leq  p \leq \hat{p} + 2 \sqrt{\frac{p(1-p)}{n}}\bigg) = 0.95
\end{eqnarray*}

We'd love to create our interval for $p$ using $\hat{p} \pm 2 \sqrt{\frac{p(1-p)}{n}}$, but we don't know $p$!  One option is to use $SE(\hat{p})$ in the estimate of the variability.

##### The Empirical Rule {-}

In a bell-shaped, symmetric distribution,

| % of data | in what interval   |
|:-------------|:----------------------------|
|$\approx 68\%$ | of the observations fall within 1 st dev of the mean |
|$\approx 95\%$ | of the observations fall within 2 st dev of the mean |
|$\approx 99.7\%$ | of the observations fall within 3 st dev of the mean |




<!--
%Show gorilla move:
%\url{viscog.beckman.uiuc.edu/flashmovie/15.php}
-->



## 1/30/20 Agenda {#Jan30}
1. Normal distribution (no q-q plots)
2. Calculating normal probabilities


### Normal Probabilities & Z scores {#norm}

#### Z score {-}

> A **Z score** of an observation is the number of standard deviations it falls above or below the mean. We compute the Z score for an observation x that follows a distribution with mean $\mu$ and standard deviation $\sigma$ using

$$ Z = \frac{x - \mu}{\sigma}$$

#### Normal probabilities {-}

We return to the Reese's Pieces example to investigate the probability of a particular number of orange candies, using the normal approximation. 

Remember:  $$SD(\hat{p}) = \sqrt{\frac{p(1-p)}{n}}$$

And the respective Z score is: $$ Z = \frac{\hat{p} - p}{\sqrt{\frac{p(1-p)}{n}}}$$

(a) What is the probability that in a sample of 25 candies, you would get less than 40% orange (provided that the machine colors 50% of the candies orange).  Answer: 0.1587

(b) What is the probability that in a sample of 250 candies, you would get less than 40% orange (provided that the machine colors 50% of the candies orange).  Answer: 0.0007888

(c) What is the probability that in a sample of 25 candies, you would get between 40% and 55% orange (provided that the machine colors 50% of the candies orange).  Answer: 0.5328


```{r}
library(mosaic)
# (a)
(0.4 - 0.5) / sqrt(0.5*0.5/25)
xpnorm(-1, 0, 1)


# (b)
(0.4 - 0.5) / sqrt(0.5*0.5/250)
xpnorm(-3.16, 0, 1)


# (c)
(0.55 - 0.5) / sqrt(0.5*0.5/25)
xpnorm(c(-1, 0.5), 0, 1)
```

Note that normal probabilities can be estimated for any variable that has a distribution which is well approximated by the bell shape given by a normal curve.  Below we calculate Z scores and probabilities for a non-proportion setting and then ask whether the values could possibly be normal.  (What do you think?)

#### Example: Athletic comparison^[taken from https://askgoodquestions.blog/2019/08/26/8-end-of-the-alphabet/ by Allan Rossman, he borrowed from 2011 AP Statistics exam]

The example below allows for a comparison between two athletes based on speed and strength.  The following information is provided about the sample of individuals who were measured:

* Speed is measured by the time required to run a distance of 40 yards, with smaller times indicating more desirable (faster) speeds.  From the data, the times to run 40 yards have a mean of 4.60 seconds and a standard devotion of 0.15 seconds, with a minimum of 4.40 seconds.

* Strength is measured by the amount of weight lifted, with more weight indicating more desirable (greater) strength  From the data, the amount of weight lifted has a mean of 310 pounds and a standard deviation of 25 pounds.


|                      | mean     | std dev  | minimum  |
|:----------------------|:----------|:----------|:----------|
| Time to run 40 yards | 4.60 sec | 0.15 sec | 4.40 sec |
| Amount lifted        | 310 lbs  | 25 lbs   | NA       |


1. Calculate and interpret the Z score for a player who can lift weight of 370 pounds.

$$Z = \frac{370-310}{25} = 2.4$$

This z-score tells us that a player who can lift 370 pounds is lifting 2.4 SDs more than average.  Saying that this weight is 2.4 SDs away from the average would leave out important information about direction.


2. Consider two players, A and B (with data given as below).  Which player should be selected for the team if only one player can be selected?

|                      | Player A | Player B |
|:----------------------|:----------|:----------|
| Time to run 40 yards | 4.42 sec | 4.57 sec |
| Amount lifted        | 370 lbs  | 375 lbs  |


At a first glance, we can see that A is faster, and B is stronger.  Understanding how each player performs (in strength and speed) relative to the rest of the players is the first step in answering the question.  We will calculate four Z scores, one for each player and each task:

\begin{align*}
Z_{Aspeed} =  \frac{4.42 - 4.6}{0.15} = -1.2\\
Z_{Astrength} = \frac{370-310}{25} = 2.4\\
Z_{Bspeed} =  \frac{4.57 - 4.6}{0.15} = -0.2\\
Z_{Bstrength} = \frac{375-310}{25} = 2.6\\
\end{align*}


After calculating Z scores, it is found that Player B is only slightly stronger than Player A, but Player A is considerably faster than Player B.  Because the question advised us to consider both criteria as equally valuable, Player A is the better choice.

3. Using the full information about the speed data, do you think that the distribution of 40 yard running times is approximately normal?

NO! The minimum is too close to the mean for the normal distribution to provide a reasonable model.  What does "too close" mean here?  Let's see how many standard deviations the minimum is below the mean:

$$ Z = \frac{4.4 - 4.6}{0.15} = -1.33 $$

The Z score tells us that the minimum speed is only -1.33 standard deviations below the mean.  According to the normal distribution (see the plot below), we would expect about 9% of the observations to be lower than 4.4 seconds, so the normal distribution does not seem to be a great fit to these observations.

```{r}
xpnorm(-1.333, 0, 1, plot = TRUE)
```

## 2/4/20 Agenda {#Feb4}
1. Theoretical basis for confidence intervals
2. $z^*$  (different from Z score!)
3. Example: extreme poverty

## Confidence Intervals {#CI}

### Theoretical set-up

#### Conditions for when the sampling distribution of $\hat{p}$ is nearly normal  (The Central Limit Theorem!!) {-} 

The sampling distribution for $\hat{p}$, taken from a sample of size $n$ from a population with a true proportion $p$, is nearly normal when:

1. the sample observations are independent
2. we expected to see at least 10 successes and 10 failures in our samples.  Said differently, $np \geq 10$ and $n(1-p) \geq 10$.  This is sometimes called the **success-failure condition**.

If the conditions are met, then the sampling distribution of $\hat{p}$ is nearly normal with mean $p$ and standard error:

$$SE_{\hat{p}} = SE (\hat{p}) = \sqrt{\frac{p(1-p)}{n}}$$

#### How far is $\hat{p}$ from $p$ ???  {-}

Great news, the $SE(\hat{p})$ measures the distance we can expect between $\hat{p}$ from $p$!!!  Indeed, a Z score tells us the distance between $\hat{p}$ from $p$ in units of standard error.

The normal distribution provides **percentages** for how often Z scores should fall in certain ranges.

From the empirical rule, we would expect 95% of our $\hat{p}$ values to be within 2 standard deviations of the mean.  That is, 95% of $\hat{p}$ are:
\begin{eqnarray*}
p \pm 2 \sqrt{\frac{p(1-p)}{n}}
\end{eqnarray*}
Or put differently, when referring to a randomly selected $\hat{p}$,
\begin{eqnarray*}
P\bigg( p - 2 \sqrt{\frac{p(1-p)}{n}} \leq \hat{p} \leq p + 2 \sqrt{\frac{p(1-p)}{n}}\bigg) = 0.95\\
P\bigg( - 2 \sqrt{\frac{p(1-p)}{n}} \leq \hat{p} - p \leq 2 \sqrt{\frac{p(1-p)}{n}}\bigg) = 0.95\\
P\bigg(\hat{p} - 2 \sqrt{\frac{p(1-p)}{n}} \leq  p \leq \hat{p} + 2 \sqrt{\frac{p(1-p)}{n}}\bigg) = 0.95
\end{eqnarray*}

Putting it all together, we create a confidence interval for $p$ which says that 95% of all samples will create confidence intervals that capture the true (unknown $p$):

$$95\% \mbox{ CI for }p:  \hat{p} \pm 1.96 \sqrt{\frac{p(1-p)}{n}}$$

And if a different percentage is needed, change the multiplier appropriately:

####  Confidence Interval Formula {-}

$$\mbox{ CI for }p:  \hat{p} \pm z^* \sqrt{\frac{p(1-p)}{n}}$$

What is $z^*$?  It is defined using the normal distribution which is centered at zero with a standard deviation of one.  

For example, if a 99% confidence interval is desired, find the $z^*$ value that captures 99% of the observations between $-z^*$ and $z^*$.

$$99\% \mbox{ CI for }p:  \hat{p} \pm 2.58 \sqrt{\frac{p(1-p)}{n}}$$

```{r}
xpnorm(c(-2.58, 2.58), 0, 1, plot = TRUE)
```


####  What does the percentage level *mean*? {-}

A *confidence level* is the long-run percent of intervals that capture the true parameter.

### Example: changes in extreme poverty

#### In-class activity set-up {-}

Recall from the in-class activity:

Some of you may be familiar with Hans Rosling who founded the website https://www.gapminder.org/ and dedicated his life to promoting awareness of global health issues, see his Ted talks here: https://www.ted.com/playlists/474/the_best_hans_rosling_talks_yo.  One question he liked to ask is:

> Has the percentage of the world’s population who live in extreme poverty doubled, halved, or remained about the same over the past twenty years?

* Before you go on, answer the question.  Has the extreme poverty doubled, halved, or remained about the same?  What do you think?


>The correct answer is that this percentage has halved, but only 5% of a sample of 1005 U.S. adults in 2017 got this right.  Rosling liked to say that chimpanzees would do better than people: With only three options, we would expect 33.33% of chimpanzees to answer correctly.

* If in fact the students are randomly guessing, how many standard deviations away from the "random guess" value is 0.05?  [Hint:  use proportions and not percentages in your calculations.]

note:  we covered this in class on Tuesday, so it's in the notes, but the formula doesn't show up in your text until the box on page 124 in section 3.1.1.

Do not use the computer here (except as a calculator, and feel free to use a calculator or use the computer / R as a calculator).  Note:  you need to know how many people were asked, look above.

**Solution**

$$SD(\hat{p}) = \sqrt{p(1-p)/n} = \sqrt{(1/3)(2/3)/1005} = 0.0149$$

```{r}
sqrt((1/3)*(2/3)/1005)
```

How far is 0.05 from (1/3) in units of standard deviation?  That's just a Z score!  Yikes, the 5% value is MORE THAN 19 STANDARD DEVIATIONS BELOW RANDOM GUESSING!!!

```{r}
Z_p = (0.05 - (1/3)) / sqrt((1/3)*(2/3)/1005)
Z_p
```


* What does this say about humans doing so much *worse* than random guessing when answering the question about poverty?  (No hypothesis test here, just a reflection on the distance between the observed data and the random guess answer.)

**Solution**

Not only are humans *wrong*, but they are *wrong* at an extremely high rate.  That is, they are wrong in such a way that they can't possibly be guessing.  There must be something about the question that makes so many people get it wrong (maybe that they are all seeing the same media narrative which describes continued problems with extreme poverty?)

We could find the percent of samples that would have produced such a small $\hat{p}$ if people were indeed random guessing.  Unsurprisingly, the proportion of such samples is exceedingly small:

```{r}
xpnorm(-19.05, 0, 1, plot=TRUE)
```


#### Confidence Interval for true population proportion

Given the extreme poverty set-up above, the question turns from one of a hypothesis test to one of a confidence interval.  Note that we are making one more change to the question, we are curious about the proportion of people who think that the rate has doubled.

\begin{eqnarray*}
p &=& \mbox{true proportion of people who incorrectly believe that the % of the}\\
&=& \mbox{ world’s population who live in extreme poverty has doubled}\\
\hat{p} &=& \mbox{sample proportion of people who incorrectly believe that the % of the}\\
&=& \mbox{ world’s population who live in extreme poverty has doubled}
\end{eqnarray*}

It turns out that in the sample of 1005 adult Americans, 593 people thought that the rate had doubled.^[see results here: https://www.gapminder.org/ignorance/gms/]

$$\hat{p} = \frac{593}{1005} = 0.59$$

A 95% confidence interval for the true proportion of adult Americans who think the rate has doubled is (0.56, 0.62).  We are 95% confident that the true proportion of adult Americans who think the extreme poverty rate has doubled is between 0.56 and 0.62.

$$ \hat{p} \pm 1.96 * \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

```{r}
593/1005 - 1.96 * sqrt((593/1005)*(412/1005) / 1005)
593/1005 + 1.96 * sqrt((593/1005)*(412/1005) / 1005)
```

Question: Survey researchers typically select only one random sample from a population, and then they produce a confidence interval based on that sample.How do we know whether the resulting confidence interval is successful in capturing the unknown value of the population parameter?  

Answer: we don't know!  We never know if the interval actually captures the parameter or not.  We just know that over our lifetime as scientists, we will capture at the rate we set.

Question: If we can’t know for sure whether the confidence interval contains the value of the population parameter, on what grounds can we be confident about this?

Answer: well, we agree about the **process** that created the CI.  

### Modifying CIs {#modCI}

#### Changing $n$ {-}

As we can see from the CI formula, increasing $n$ has the effect of **decreasing** the width of the CI.  

$$ \hat{p} \pm 1.96 * \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

#### Changing $p$ {-}

A different value of $p$ means that the sampling distribution will have a different center (and a different SE), but the coverage rate will not change, and the SE probably won't change very much.

#### Changing the confidence level {-}

The choice of $z^*$ determines (over, say, your lifetime as a scientist) the percent of your research confidence intervals that will capture the true parameter of interest.  Note that the larger the $z^*$ value, the more likely it is that a sample will produce a CI which captures the true parameter.

Note that $z^* = 1.645$ produces CIs that capture at a 90% rate.  $z^* = 2.58$ produces CIs that capture at a 99% rate.

```{r}
xpnorm(1.645, 0, 1)
xpnorm(2.58, 0, 1)
```

Question:  why don't we always use 99.99% CIs?

Answer:  because the intervals would typically be too wide to provide any real information about the actual population parameter.

## 2/6/20 Agenda {#Feb6}
1. Putting together all the pieces of the CI
2. Effects of sample size, $p$, and confidence level on CI
3. What is the confidence level?



## 2/11/20 Agenda {#Feb11}
1. Simple Random Sampling
2. Biased sampling

## Sampling {#samp}

Suppose that an alien lands on earth and wants to estimate the proportion of human beings who are female.  Fortunately, the alien took a good statistics course on its home planet, so it knows to take a sample of human beings and produce a confidence interval for this proportion.  Unfortunately, the alien happens upon the 2019 U.S. Senate as its sample of human beings.  The U.S. Senate has 25 women senators (its most ever!) among its 100 members in 2019.

a) Calculate the alien’s 95% confidence interval.  

This calculation becomes .25 ± .085, which is the interval (.165 → .335).

b) Interpret the interval.  The alien would be 95% confident that the proportion of all humans on earth who are female is between .165 and .335.

c) Is this consistent with your experience living on this planet?  No, the actual proportion of humans who are female is much larger than this interval, close to 0.5.

d) What went wrong?  The alien did not select a random sample of humans.  In fact, the alien’s sampling method was very biased toward under-representing females.

e) As we saw with the applet, about 5% of all 95% confidence intervals fail to capture the actual value of the population parameter.  Is that the explanation for what went wrong here?  No!  Many students are tempted to answer yes, but this explanation about 5% of all intervals failing is only relevant when you have selected random samples over and over again.  The lack of random sampling is the problem here.

f) Would it be reasonable for the alien to conclude, with 95% confidence, that between 16.5% and 33.5% of U.S. senators in the year 2019 are female?  No.  We know (for sure, with 100% confidence) that exactly 25% of U.S. senators in 2019 are female.  If that’s the entire population of interest, there’s no reason to calculate a confidence interval.  This question is a very challenging one, for which most students need a nudge in the right direction.

The lessons of this example are:

Confidence intervals are not appropriate when the data were collected with a biased sampling method.  A confidence interval calculated from such a sample can provide very dubious and misleading information.
Confidence intervals are not appropriate when you have access to the entire population of interest.  In this unusual and happy circumstance, you should simply describe the population.

## 2/13/20 Agenda {#Feb13}
1. Type I & Type II errors
2. Power

## Errors & Power {#errors}

