--- 
title: "Introduction to (Bio)Statistics"
author: "Jo Hardin"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
header-includes:  \usepackage{blkarray}
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: hardin47/website/Math58/
description: "Class notes for both Math 58 and Math 58B at Pomona College: Introduction to Statistics and Introduction to Biostatistics.  The notes are based extensively Introductory Statistics with Randomization and Simulation by Diez, Barr, and Cetinkaya-Rundel."
---


# Class Information {-}

Class notes for Math 58(B) at Pomona College: Introduction to (Bio)Statistics.  The notes are based extensively on "Introductory Statistics with Randomization and Simulation" by Diez, Barr, and \c{C}etinkaya-Rundel [@isrs].  We will also be using many examples (and applets) from "Investigating Statistics, Concepts, Applications, and Methods" by Chance and Rossman [@iscam].


You are responsible for reading the relevant chapters in the text.  The text is very good & readable, so you should use it.  Also, you will have a much deeper understanding of the material if you spend time working through the applets at http://www.rossmanchance.com/iscam3/files.html.  You should make sure you are coming to class and also reading the materials associated with the activities. 

```{r eval=FALSE, include = FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")

setwd("~/gitRepos/website/Math58")
#bookdown::render_book("index.Rmd", "bookdown::gitbook")

bookdown::render_book("index.Rmd", output_dir = "Notes")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```



```{r include = FALSE, eval=FALSE}
# iscam
# http://www.rossmanchance.com/iscam3/files.html

# openintro
#https://www.openintro.org/

# modern dive: https://moderndive.com/
```


```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, 
                      fig.align = "center")

library(tidyverse)
library(broom)
library(caret)
library(infer)
```

| Day    	| Topic     	|  Book Chap   	|   Notes Section |
|:-------	|:------------|:---------	|:--------------------	|
| 1/21/20 	| Intro to Data / R |  | \@ref(intro) [intro],  \@ref(Jan21) [Jan21]|
| 1/23/20	| Foundations for Inference | ISRS 2.1 - 2.3 |  \@ref(Jan23) [Jan23], \@ref()  [] |
