--- 
title: "Linear Models"
author: "Jo Hardin"
date: "`r Sys.Date()`"
knit: bookdown::render_book
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: hardin47/website/Math158/
description: "Class notes for Math 158 at Pomona College: Linear Models.  The notes are based extensively on An Introduction to Statistical Learning by James, Witten, Hastie, and Tibshirani as well as  Applied Linear Statistical Models, 5th edition by Kutner, Nachtsheim, Neter, Li."
---


# Class Information {-}

Class notes for Math 158 at Pomona College: Linear Models.  The notes are based extensively on An Introduction to Statistical Learning [@ISL];  Applied Linear Statistical Models [@kutner].  The computing part of the class will be taken from R for Data Science [@r4ds] by Wickham and Grolemund as well as Tidy Modeling with R [@tidymodels] by Kuhn and Silge.


You are responsible for reading the relevant chapters in the texts.  The texts are very good & readable, so you should use them.   You should make sure you are coming to class and also reading the materials associated with the activities. 

```{r eval=FALSE, include = FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")

setwd("~/gitRepos/web stuff/website/Math158")
#bookdown::render_book("index.Rmd", "bookdown::gitbook")

bookdown::render_book("index.Rmd", output_dir = "Notes")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```



```{r include = FALSE, eval=FALSE}
# http://personal.psu.edu/drh20/525/
# http://personal.psu.edu/drh20/525/assignments/

# categorical & poisson: https://newonlinecourses.science.psu.edu/statprogram/stat504
# https://newonlinecourses.science.psu.edu/stat504/

# broadening your statistical horizons (roback & legler) https://bookdown.org/roback/bookdown-bysh/

# modern dive: https://moderndive.com/

```


```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, 
                      fig.align = "center")

library(tidyverse)
library(broom)
library(caret)
```

```{r echo = FALSE, eval = FALSE}
# from 2018
| Day    	| Topic     	|  Book Chap   	|   Notes Section |
|:-------	|:------------|:---------	|:--------------------	|
1/15 & Intro \& SLR & 1 (ALSM)  \& 3.1 (ISLR)\\
\\
1/22 &  Inference in Regression & 2  (ALSM) \\
\\
1/29 &Diagnostics &  3  (ALSM)  \\
\\
2/5 & Simultaneous Inference \& Matrices & 4 \& 5 (ALSM) \\
\\
2/12 & Multiple Regression I & 6  (ALSM) \& 3.2 (ISLR) \\
\\
2/19 & MLR Catch-up  & \\
{\bf 2/21} & {\bf Wednesday, Exam 1} & 1-6 (ASLM) \\
\\
2/26 & Multiple Regression & 7 (ALSM)\\
\\
3/5 & Model Building I \& II & 9  (ALSM) / 6.1 (ISLR) \& 10 (ALSM)  \\
\\
3/12 & Spring Break &  \\
\\
3/19 & Shrinkage & 6.2 (ISLR) \\
\\
3/26 & Smoothing Splines & 7 (ISLR) \\
\\
4/2 &  Kernel Smoothers & \\
\\
4/9 & Catch-up &\\
{\bf 4/11} & {\bf Wednesday, Exam 2} & 7-10 (ALSM) \& ISLR\\
\\
4/16 & ANOVA & 16/17 (ALSM)\\
\\
4/23 & two-way ANOVA / diagnostics & 19/18 (ALSM)\\
\\
4/30 &  fun of ANOVA Catch-up & \\
{\bf 5/2} & {\bf Wednesday, Exam 3} & \\
\\
{\bf 5/9} & {\bf Wednesday 5pm, Take Home Project Due} & \\
```
