# Simple Linear Regression

more soon



## R Example (SLR): Happy Planet  
  

The data below represents 10 different variables on health of a country measured on 143 countries.  Data taken from [@Lock5], originally from the Happy Planet Index Project [http://www.happyplanetindex.org/].  Region of the world is coded as 1 = Latin America, 2 = Western nations, 3 = Middle East, 4 = Sub-Saharan Africa, 5 = South Asia, 6 = East Asia, 7 = former Communist countries.  We are going to investigate happiness and life expectancy.  

```{r warning=FALSE, message=FALSE, include=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

### Reading the data into R
```{r warning=FALSE, message=FALSE}
happy <- read_delim("~/Dropbox/teaching/math150/spring17/happyPlanet.txt", delim="\t")
glimpse(happy)  
```


### Running the linear model (lm)

```{r}
happy.lm = lm(LifeExpectancy ~ Happiness, data=happy) 

happy.lm %>% tidy()
```

### Ouptut

Some analyses will need the residuals, fitted values, or coefficients individually.
```{r}
happy.lm %>% augment()
```


We can plot the main relationship, or we can plot the residuals (to check that technical conditions hold):
```
ggplot(happy, aes(x=Happiness, y=LifeExpectancy)) + geom_point() + 
         geom_smooth(method="lm", se=FALSE) 
happy.lm %>% augment %>% ggplot(aes(x = .fitted, y = .resid)) + geom_point() + 
         geom_hline(yintercept=0)
```

```{r fig.height=3, echo=FALSE}
p1 <- ggplot(happy, aes(x=Happiness, y=LifeExpectancy)) + geom_point() + 
         geom_smooth(method="lm", se=FALSE) 
p2 <- happy.lm %>% augment %>% ggplot(aes(x = .fitted, y = .resid)) + geom_point() +
         geom_hline(yintercept=0)
multiplot(p1, p2, cols=2)
```


Intervals of interest: mean response, individual response, and parameter(s).
```{r}
predict.lm(happy.lm, newdata=list(Happiness=c(4,7)),interval=c("conf"), level=.95)
predict.lm(happy.lm, newdata=list(Happiness=c(4,7)),interval=c("pred"), level=.95)

happy.lm %>% tidy(conf.int = TRUE)
```
  
  
  
  





