# Text analysis {#sec-text-analysis}


```{r}
#| include: false

source("_common.R")
fontawesome::fa_html_dependency()
```




## Variable Types

Some new variable types: 

* character strings
* factor variables
* dates
* numeric (decimal)
* integer
* logical (Boolean)

A variable's type determines the values that the variable can take on and the operations that can be performed on it.  Specifying variable types ensures the dataset's integrity and increases performance. 



## Character strings

When working with character strings, we might want to detect, replace, or extract certain patterns. 

Strings are objects of the **character** class (abbreviated as `<chr>` in tibbles). When you print out strings, they display with double quotes:

```{r}
some_string <- "banana"
some_string
```

### Creating strings

Creating strings by hand is useful for testing out regular expressions.

To create a string, type any text in either double quotes `"` or single quotes `'`. Using double or single quotes doesn't matter unless your string itself has single or double quotes.

```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'

string1
string2
```

### Working with `str_*()` functions

#### `str_view()`

We can view these strings more "naturally" (without the opening and closing quotes) `with str_view()`:

```{r}
str_view(string1)
str_view(string2)
```

#### `str_c`

Similar to `paste()` (gluing strings together), but works well in a tidy pipeline.

```{r}
df <- tibble(name = c("Flora", "David", "Terra", NA))
df |> mutate(greeting = str_c("Hi ", name, "!"))
```

#### `str_sub()`

`str_sub(string, start, end)` will extract parts of a `string` where `start` and `end` are the positions where the substring starts ane ends.

```{r}
fruits <- c("Apple", "Banana", "Pear")
str_sub(fruits, 1, 3)
str_sub(fruits, -3, -1)
```


Won't fail if the string is too short.

```{r}
str_sub(fruits, 1, 5)
```

#### `str_sub()` in a pipeline

We can use the `str_*()` functions inside the `mutate()` function.

```{r}
#| echo: false
titanic <- Titanic |>
  as.data.frame()
```

```{r}
titanic |>
  mutate(class1 = str_sub(Class, 1, 1))
```


#### `str_replace*()`

`str_replace()` replaces the first match of a pattern. `str_replace_all()` replaces all the matches of a pattern.

```{r}
fruits
str_replace(fruits, "a", "x")
str_replace_all(fruits, "a", "x")
```

#### `str_detect()` 


```{r}
str_detect(fruits, "a")
```


`str_detect()` can be seamlessly used in a `filter()` pipeline.

::: {.panel-tabset}

## original data

```{r}
starwars |>
  select(name, films)
```

## unnested data

```{r}
starwars |>
  select(name, films) |>
  unnest_wider(films, names_sep = "")
```

## filtered unnested data

```{r}
starwars |>
  filter(str_detect(films, "Empire")) |>
  select(name, films) |>
  unnest_wider(films, names_sep = "")
```
:::

#### **stringr** functions

The **stringr** package within tidyverse contains lots of functions to help process strings. Letting x be a string variable...

```{r}
#| echo: false

tribble(
  ~`str function`       , ~arguments                      , ~returns                    ,
  "`str_vew()`"         , "`x`"                           , "the string"                ,
  "`str_c()`"           , " ..., `sep`, `collapse`"       , "a new concatenated string" ,
  "`str_sub()`"         , "`x`, `start`, `end`"           , "a modified string"         ,
  "`str_replace()`"     , "`x`, `pattern`, `replacement`" , "a modified string"         ,
  "`str_replace_all()`" , "`x`, `pattern`, `replacement`" , "a modified string"         ,
  "`str_detect()`"      , "`x`, `pattern`"                , "TRUE/FALSE"                ,
  "`str_to_lower()`"    , "`x`"                           , "a modified string"         ,
  "`str_to_upper()`"    , "`x`"                           , "a modified string"         ,
  "`str_length()`"      , "`x`"                           , "a number"                  ,
) |>
  kableExtra::kable()
```

Use the <a href = "https://opensource.posit.co/resources/cheatsheets/strings/" target = "_blank">**stringr** cheatsheet</a>. 

## Factor variables

Factor variables are a special type of character string.
The computer actually stores them as *integers* (?!?!!?) (abbreviated as `<fct>` in tibbles). 

* categorical variable
* represented in discrete levels

### Order matters

SurveyUSA poll from 2012 on views of the DREAM Act.

What is off about the data viz part of the report?

::: {.panel-tabset}

## Data
```{r}
openintro::dream
```

## Plot
```{r}
dream |>
  ggplot(aes(x = ideology, fill = stance)) +
  geom_bar()
```

## Levels
```{r}
dream |>
  select(ideology) |>
  pull() |> # because levels() works only on vectors, not data frames
  levels()
```

:::

#### Change the order

We can fix the order of the `ideology` variable.
The function `fct_relevel()` is in the **forcats** pacakge.

::: {.panel-tabset}

## Code
```{r}
#| eval: false
dream |>
  mutate(
    ideology = fct_relevel(ideology, c("Liberal", "Moderate", "Conservative"))
  ) |>
  ggplot(aes(x = ideology, fill = stance)) +
  geom_bar()
```

## Plot
```{r}
#| echo: false
dream |>
  mutate(
    ideology = fct_relevel(ideology, c("Liberal", "Moderate", "Conservative"))
  ) |>
  ggplot(aes(x = ideology, fill = stance)) +
  geom_bar()
```
:::

#### Factor *and* character variables


::: {.panel-tabset}

## Data
```{r}
starbucks |>
  select(item, type, calories)
```

## Plot
```{r}
#| echo: false
starbucks |>
  ggplot(aes(x = type, y = calories)) +
  geom_point() +
  labs(
    x = "type of food",
    y = "",
    title = "Calories for food items at Starbucks"
  )
```
:::


### Reorder according to another variable

Lets say that we wanted to order the type of food item based on the average number of calories in that food.


::: {.panel-tabset}

## Code
```{r}
#| eval: false
starbucks |>
  mutate(type = fct_reorder(type, calories, .fun = "mean", .desc = TRUE)) |>
  ggplot(aes(x = type, y = calories)) +
  geom_point() +
  labs(
    x = "type of food",
    y = "",
    title = "Calories for food items at Starbucks"
  )
```

## Plot
```{r}
#| echo: false
starbucks |>
  mutate(type = fct_reorder(type, calories, .fun = "mean", .desc = TRUE)) |>
  ggplot(aes(x = type, y = calories)) +
  geom_point() +
  labs(
    x = "type of food",
    y = "",
    title = "Calories for food items at Starbucks"
  )
```

:::


#### **forcats** functions

The **forcats** package within tidyverse contains lots of functions to help process factor variables Use the <a href = "https://opensource.posit.co/resources/cheatsheets/factors/" target = "_blank">**forcats** cheatsheet</a>. We'll focus on the most common functions. 

* functions for changing the **order** of factor levels
    - `fct_relevel()` = *manually* reorder levels
    - `fct_reorder()` = reorder levels according to values of another *variable*
    - `fct_infreq()` = order levels from highest to lowest frequency
    - `fct_rev()` = reverse the current order

* functions for changing the **labels** or values of factor levels
    - `fct_recode()` = *manually* change levels
    - `fct_lump()` = *group together*  least common levels



## Time and Date


### Working with time and date

The (very well named) R package **lubridate** is used for wrangling time and date objects [@lubridate].  In particular, **lubridate** makes it very easy to work with days, times, and dates.  The base idea is to start with dates in a `ymd` (year month day) format and transform the information into whatever you want.   The <a href = "https://opensource.posit.co/resources/cheatsheets/lubridate/" target = "_blank">**lubridate** cheatsheet</a> provides many of the basic functionality.

Example from  https://lubridate.tidyverse.org/reference/lubridate-package.html

### If anyone drove a time machine, they would crash

The length of months and years change so often that doing arithmetic with them can be unintuitive. Consider a simple operation, January 31st + one month. Should the answer be:


1. February 31st (which doesn't exist)
2. March 4th (31 days after January 31), or
3. February 28th (assuming its not a leap year)

A basic property of arithmetic is that a + b - b = a. Only solution 1 obeys the mathematical property, but it is an invalid date. Wickham wants to  make **lubridate** as consistent as possible by invoking the following rule: if adding or subtracting a month or a year creates an invalid date, **lubridate** will return an NA. 

If you thought solution 2 or 3 was more useful, no problem. You can still get those results with clever arithmetic, or by using the special `%m+%` and `%m-%` operators. `%m+%` and `%m-%` automatically roll dates back to the last day of the month, should that be necessary.

### R examples

Some basics in **lubridate**.


```{r}
require(lubridate)
rightnow <- now()

day(rightnow)
week(rightnow)
month(rightnow, label = FALSE)
month(rightnow, label = TRUE)
year(rightnow)

minute(rightnow)
hour(rightnow)

yday(rightnow)
mday(rightnow)
wday(rightnow, label = FALSE)
wday(rightnow, label = TRUE)
```

But how do I create a date object?

```{r}
jan31 <- ymd("2021-01-31")
jan31 + months(0:11)

floor_date(jan31, "month") + months(0:11) + days(31)
jan31 + months(0:11) + days(31)

jan31 %m+% months(0:11)
```

NYC flights

```{r}
library(nycflights13)
names(flights)
flightsWK <- flights |>
  mutate(ymdday = ymd(paste(year, month, day, sep = "-"))) |>
  mutate(weekdy = wday(ymdday, label = TRUE), whichweek = week(ymdday))

head(flightsWK)

flightsWK <- flights |>
  mutate(ymdday = ymd(paste(year, "-", month, "-", day))) |>
  mutate(weekdy = wday(ymdday, label = TRUE), whichweek = week(ymdday))

flightsWK |>
  select(
    year,
    month,
    day,
    ymdday,
    weekdy,
    whichweek,
    dep_time,
    arr_time,
    air_time
  ) |>
  head()
```



## <i class="fas fa-lightbulb"></i> Reflection questions  

* What is the difference between character strings, factor variables, dates, numeric variables, integer variables, and logical variables?
* What are the different `str_*()` functions? What do they do?
* What are the different `fct_*()` functions? What do they do?
* What does the `all` do in functions like `str_replace_all()` (as compared with `str_replace()`)?
* Are the `str_*()` functions case sensitive?
* Can `str_*()` functions take variables that aren't strings?
* Can `fct_*()` functions take variables that aren't factors?
* Can **lubridate** functions take variables that aren't time or data objects?
* How do the `str_*()` functions work inside the tidy pipeline?

## <i class="fas fa-balance-scale"></i> Ethics considerations 

* Why is the calculation of `January 31 + one month` interesting and/or an ethical consideration?
* Why is it often important to relevel your factor variables when making a data visualization?
* Why are timezones important to pay attention to?
* What does this mean: If x is a string it can take any value. If x is a factor it can only take a values from a list of all levels. ?
* Name one thing that you noticed about types of variables in the course materials (either in class or reading the notes, etc.) where you thought to yourself "Oh, I'll have to be really careful about **that**."  Why would you need to be careful?


## References

