# Writing good code {#sec-good-code}


```{r, include=FALSE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.height = 3,
  fig.width = 5,
  cache = TRUE,
  fig.align = "center"
)
```

```{r}
#| include: false

source("_common.R")
fontawesome::fa_html_dependency()
```



## Code style

> Good coding style is like correct punctuation: you can manage without it, butitsuremakesthingseasiertoread.^[https://style.tidyverse.org/]

All of the following examples are taken from the <a href = "https://style.tidyverse.org/" target = "_blank">Tidyverse style guide</a>.

### Object names

> There are only two hard things in Computer Science: cache invalidation and naming things.

> -- Phil Karlton


Variable and function names should use only lowercase letters, numbers, and `_`. Use underscores (`_`) (so called snake case) to separate words within a name.

```{r}
#| eval: false
# Good
day_one
day_1

# Bad
DayOne
dayone

# Really bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```


### Spacing

Do not put spaces inside or outside parentheses for regular function calls.

```{r}
#| eval: false
# Good
mean(x, na.rm = TRUE)

# Bad
mean(x, na.rm = TRUE)
mean(x, na.rm = TRUE)
```


### Infix operators

Most infix operators (`==`, `+`, `-`, `<-`, etc.) should always be surrounded by spaces:

```{r}
#| eval: false
# Good
height <- (feet * 12) + inches
mean(x, na.rm = TRUE)

# Bad
height <- feet * 12 + inches
mean(x, na.rm = TRUE)
```

Fun note: many languages have infixes that naturally change the meaning of a word. In English we have many prefixes and suffixes, for example, unhappy ("un" is the prefix) or hopeless ("less" is the suffix). There is only one infix in the English language: "friggin" (and its derivatives), as in: unfrigginbelievable.

### Long function calls

If a function call is too long to fit on a single line, use one line each for the function name, each argument, and the closing ). This makes the code easier to read and to change later.

```{r}
#| eval: false
# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated(
  "that",
  requires,
  many,
  arguments,
  "some of which may be long"
)
```


### Long lines (piping)

If the arguments to a function don't all fit on one line, put each argument on its own line and indent:^[The last one isn't great because it is hard to distinguish the data frame from the other arguments (which play very different roles in the `summarize()` function).]

```{r}
#| eval: false
# Good
iris |>
  summarise(
    Sepal.Length = mean(Sepal.Length),
    Sepal.Width = mean(Sepal.Width),
    .by = Species
  )

# Bad
iris |>
  summarise(
    Sepal.Length = mean(Sepal.Length),
    Sepal.Width = mean(Sepal.Width),
    .by = Species
  )

# Also bad
summarise(
  iris,
  Sepal.Length = mean(Sepal.Length),
  Sepal.Width = mean(Sepal.Width),
  .by = Species
)
```

### Short lines (piping)

Sometimes it's useful to include a short pipe as an argument to a function in a longer pipe. Carefully consider whether the code is more readable with a short inline pipe or if it's better to move the code outside the pipe and give it an evocative name.

```{r}
#| eval: false
# Good
x |>
  semi_join(y |> filter(is_valid))

# Ok
x |>
  select(a, b, w) |>
  left_join(y |> select(a, b, v), join_by(a, b))

# Better
x_join <- x |> select(a, b, w)
y_join <- y |> select(a, b, v)
left_join(x_join, y_join, join_by(a, b))
```

### Style

* The point is that coding happens in community.  
* Not only do you want your code to run well, but you want other people to be able to understand it and use it. 
* The more that you and others use the same syntax, the better the communication will be.





## <i class="fas fa-lightbulb"></i> Reflection questions  

* What are some of the general rules for writing clear code?

* Give one reason why piping into a data verb is preferred to using the data frame as the first argument.

## <i class="fas fa-balance-scale"></i> Ethics considerations 

* What could go wrong if your code style doesn't match what is expected by your collaborator? Or if it is hard to read by your collaborator?

* Why is good communication important in data science?


## References
