--- 
title: "Computational Statistics"
author: "Jo Hardin"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: hardin47/website/Math154/
description: "Class notes for Math 154 at Pomona College: Computational Statistics.  The notes are based extensively on An Introduction to Statistical Learning by James, Witten, Hastie, and Tibshirani as well as Modern Data Science with R by Baumer, Kaplan, and Horton."
---


# Class Information {-}

Class notes for Math 154 at Pomona College: Methods in Biostatistics.  The notes are based extensively on An Introduction to Statistical Learning [@ISL] by James, Witten, Hastie, and Tibshirani as well as Modern Data Science [@MDSR] with R by Baumer, Kaplan, and Horton.


You are responsible for reading the relevant chapters in the text.  The texts are very good & readable, so you should use them.   You should make sure you are coming to class and also reading the materials associated with the activities. 

```{r eval=FALSE, include = FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")

setwd("~/gitRepos/website/Math154")
#bookdown::render_book("index.Rmd", "bookdown::gitbook")

bookdown::render_book("index.Rmd", output_dir = "Notes")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```



```{r include = FALSE, eval=FALSE}
# http://personal.psu.edu/drh20/525/
# http://personal.psu.edu/drh20/525/assignments/

# categorical & poisson: https://newonlinecourses.science.psu.edu/statprogram/stat504
# https://newonlinecourses.science.psu.edu/stat504/

# broadening your statistical horizons (roback & legler) https://bookdown.org/roback/bookdown-bysh/

# modern dive: https://moderndive.com/

```


```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, 
                      fig.align = "center")

library(tidyverse)
library(broom)
library(caret)
```

| Day    	| Topic                      	|  Book Chp   	|   Notes Section |
|:---------	|:---------------------------	|:----------------:	|:----------------:	|
| 1/23/19 	| t-tests / SLR / Intro to R | 2 | \@ref(intro), \@ref(ttest), \@ref(tslr) [t-test] |
| 1/28/19 	| SLR | 2 |  \@ref(SLR)  [SLR]   |
| 1/30/19 	| | | \@ref(intervals), \@ref(cat) [SLR CI] |
| 2/4/19  	| Contingency Analysis | 6 | \@ref(fisher), \@ref(catest)   [.Fisher's Exact Test] |
| 2/6/19 	| | |  \@ref(studies) [Types of studies]
| 2/11/19 	| Contingency Analysis | 6 | \@ref(catest) [RR and OR] |

