--- 
title: "Computational Statistics"
author: "Jo Hardin"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::pdf_book:
    includes:
      in_header: preamble.tex
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: hardin47/website/Math154/
description: "Class notes for Math 154 at Pomona College: Computational Statistics.  The notes are based extensively on An Introduction to Statistical Learning by James, Witten, Hastie, and Tibshirani as well as Modern Data Science with R by Baumer, Kaplan, and Horton."
---


# Class Information {-}

Class notes for Math 154 at Pomona College: Methods in Biostatistics.  The notes are based extensively on An Introduction to Statistical Learning [@ISL] by James, Witten, Hastie, and Tibshiran;  Modern Data Science [@MDSR] with R by Baumer, Kaplan, and Horton; Data Science in R: A Case Studies Approach to Computational Reasoning and Problem Solving [@DSR] by Nolan and Temple Lang; and Visual and Statistical Thinking: Displays of Evidence for Making Decisions [@VST] by Tufte.


You are responsible for reading the relevant chapters in the text.  The texts are very good & readable, so you should use them.   You should make sure you are coming to class and also reading the materials associated with the activities. 

```{r eval=FALSE, include = FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")

setwd("~/gitRepos/website/Math154")
#bookdown::render_book("index.Rmd", "bookdown::gitbook")

bookdown::render_book("index.Rmd", output_dir = "Notes")
```


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```



```{r include = FALSE, eval=FALSE}
# http://personal.psu.edu/drh20/525/
# http://personal.psu.edu/drh20/525/assignments/

# categorical & poisson: https://newonlinecourses.science.psu.edu/statprogram/stat504
# https://newonlinecourses.science.psu.edu/stat504/

# broadening your statistical horizons (roback & legler) https://bookdown.org/roback/bookdown-bysh/

# modern dive: https://moderndive.com/

```


```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, 
                      fig.align = "center")

library(tidyverse)
library(broom)
library(caret)
```

| Day    	| Topic     	|  Book Chap   	|   Notes Section |
|:-------	|:------------|:---------:	|:--------------------	|
| 9/3/19 	| Intro to Data / R | ISL1 | \@ref(intro) [intro],  \@ref(Sep3) [Sep3]|
| 9/5/19	| Intro to GitHub |  |  \@ref(Sep5) [Sep5], \@ref(repro)  [repro] |
| 9/10/19	| Data Viz| VST | \@ref(Sep10) [Sep10], \@ref(thoughts) [thoughts]  |
| 9/12/19  	|  | MDS2 |  \@ref(Sep12)   [Sep12], \@ref(deconstruct) [decon], \@ref(gg) [gg]|
| 9/17/19	| Data Wrangling | [MDS4](http://mdsr-book.github.io/) | \@ref(Sep17) [Sep17], \@ref(datastruc) [structure] |
| 9/19/19  	|  |  | \@ref(Sep17) [Sep17], \@ref(highverb)   [highverb] |
| 9/24/19	| Simulating |  MDS8 | \@ref(Sep24) [Sep24], \@ref(simmodels) [sim models] |
| 9/26/19  	|  |  | \@ref(Sep26) [Sep26], \@ref(simsens)   [sim sens] |
| 10/1/19	| Permutations |  [ISCAM](http://www.rossmanchance.com/iscam3/files.html) | \@ref(Oct1) [Oct1], \@ref(algs) [inf algs] |
| 10/3/19  	|  |  | \@ref(Oct3) [Oct3], \@ref(perms)   [perm test] |




